---
title: "One billion row challenge using base R"
author:
  - name: David Schoch
    orcid: 0000-0003-2952-4812
date: 2024-01-08
categories: [R]
---

One of my new years resolutions is to blog a bit more on the random shenanigans I do with R. This is one of those.

The [One Billion Row challenge](https://www.morling.dev/blog/one-billion-row-challenge/) by Gunnar Morling is as follows:

> write a Java program for retrieving temperature measurement values from a text file and calculating the min, mean, and max temperature per weather station. There’s just one caveat: the file has 1,000,000,000 rows!

I didn't take long, also thanks to [Hacker News](https://news.ycombinator.com/item?id=38851337), that the challenge spread to other programming languages. The original repository contains a [show & tell](https://github.com/gunnarmorling/1brc/discussions/categories/show-and-tell) where other results can be discussed.

Obviously it also spread to R and there is a [GitHub repository](https://github.com/alejandrohagan/1br) from Alejandro Hagan dedicated to the challenge. There were some [critical discussions](https://github.com/alejandrohagan/1br/issues/5) on the seemingly bad performance of `data.table` but that issue thread also evolved to a discussion on other solutions. 

The obvious candidates for fast solutions with R are [`dplyr`](https://github.com/tidyverse/dplyr), [`data.table`](https://github.com/Rdatatable/data.table), [`collapse`](https://github.com/SebKrantz/collapse), and [`polars`](https://github.com/pola-rs/r-polars). From those, it appears that polars might solve the tasks the [fastest](https://github.com/alejandrohagan/1br/issues/5#issuecomment-1879737918).

I was curious, how far one can get with base R. 

## Creating the data

The R repository contains a [script](https://github.com/alejandrohagan/1br/blob/main/generate_data.R) to generate benchmark data. For the purpose of this post, I created files with 1e6 and 1e8 rows. Unfortunately, my personal laptop cannot handle 1 billion rows without dying.

## Reading the data

All base R functions will profit from reading the station column as a factor instead of a usual string. How would have thought that `stringAsFactors = TRUE` can be useful.

```{r}
#| label: read_1e6
D <- data.table::fread("measurements1e6.csv", stringsAsFactors = TRUE)
D
```

## The obvious: aggregate and split/lapply

The most obvious choice for me was to use `aggregate()`.

```{r}
#| label: aggregate
sum_stats_vec <- function(x) c(min = min(x), max = max(x), mean = mean(x))
aggregate(measurement ~ state, data = D, FUN = sum_stats_vec) |> head()
```

I was pretty sure that this might be the best solution.

The other obvious solution is to split the data frame according to staty and then `lapply` the stats calculation on each list entry

```{r}
#| label: split
split_lapply <- function(D) {
    result <- lapply(split(D, D$state), function(x) {
        stats <- sum_stats_vec(x$measurement)
        data.frame(
            state = unique(x$state),
            min = stats[1],
            max = stats[2],
            mean = stats[3]
        )
    })
    do.call("rbind", result)
}
split_lapply(D) |> head()
```

## The elegant: by

I stumbled upon `by` when searching for alternatives. I think it is a quite elegant looking way of solving a group/summarize task with base R. Unfortunately it returns a list and not a data frame or matrix (I made that an implicit requirement).  

In the help for `by` I stumbled upon a function I wasn't aware of yet: `array2DF`! 
```{r}
#| label: by
array2DF(by(D$measurement, D$state, sum_stats_vec)) |> head()
```

Does exactly what is needed here. For the benchmarks, I will also include a version without the `array2DF` call, to check its overhead.

## Another apply: tapply

In the help for `by`, I also stumbled upon this sentence 

> Function `by` is an object-oriented wrapper for `tapply` applied to data frames.

So maybe we can construct a solution that uses tapply, but without any inbuilt overhead in `by`.

```{r}
#| label: tapply1
do.call("rbind", tapply(D$measurement, D$state, sum_stats_vec)) |> head()
```

At this point, I was also curious if the `do.call("rbind",list)` can be sped up, so I constructed a second tapply solution.

```{r}
#| label: tapply2
sapply(tapply(D$measurement, D$state, sum_stats_vec), rbind) |> head()
```

## The obscure: reduce

I thought that this should be it, but the I remembered `reduce` exists.

```{r}
#| label: reduce

reduce <- function(D) {
    state_list <- split(D$measurement, D$state)
    Reduce(function(x, y) {
        res <- sum_stats_vec(state_list[[y]])
        rbind(x, c(state = y, mean = res[1], min = res[2], max = res[3]))
    }, names(state_list), init = NULL)
}

reduce(D) |> head()
```

technically, this is a bit wrong since it it converts the numerical values to characters. But I allow it for this experiment.

## The unfair contender: Rfast

Pondering about how this functions could be sped up in general, I remembered the package [`Rfast`](https://github.com/RfastOfficial/Rfast/) and managed to construct a solution using this package.

```{r}
#| label: Rfast

Rfast <- function(D) {
    lev_int <- as.numeric(D$state)
    minmax <- Rfast::group(D$measurement, lev_int, method = "min.max")
    data.frame(
        state = levels(D$state),
        mean = Rfast::group(D$measurement, lev_int, method = "mean"),
        min = minmax[1, ],
        max = minmax[2, ]
    )
}

Rfast(D) |> head()
```

Pretty sure that this will be the fastest, maybe even competitive with the other big packages!

## Benchmark

To reorder the benchmark results from `microbenchmark`, we use a function provided by [Dirk Eddelbuettel](https://github.com/eddelbuettel/dang/blob/master/R/reorderMicrobenchmarkResults.R).
```{r}
#| label: bench_print
reorderMicrobenchmarkResults <- function(res, order = "median") {
    stopifnot("Argument 'res' must be a 'microbenchmark' result" = inherits(res, "microbenchmark"))

    smry <- summary(res)
    res$expr <- factor(res$expr,
        levels = levels(res$expr)[order(smry[["median"]])],
        ordered = TRUE
    )
    res
}
```


```{r}
#| label: bench-1e6
#| cache: true
sum_stats_list <- function(x) list(min = min(x), max = max(x), mean = mean(x))
sum_stats_tibble <- function(x) tibble::tibble(min = min(x), max = max(x), mean = mean(x))

bench1e6 <- microbenchmark::microbenchmark(
    aggregate = aggregate(measurement ~ state, data = D, FUN = sum_stats_vec),
    split_lapply = split_lapply(D),
    array2DF_by = array2DF(by(D$measurement, D$state, sum_stats_vec)),
    raw_by = by(D$measurement, D$state, sum_stats_vec),
    docall_tapply = do.call("rbind", tapply(D$measurement, D$state, sum_stats_vec)),
    sapply_tapply = sapply(tapply(D$measurement, D$state, sum_stats_vec), rbind),
    reduce = reduce(D),
    Rfast = Rfast(D),
    dplyr = D |> dplyr::group_by(state) |> dplyr::summarise(sum_stats_tibble(measurement)) |> dplyr::ungroup(),
    datatable = D[, .(sum_stats_list(measurement)), by = state],
    times = 50
)
```

```{r}
#| label: fig-bench1e6
ggplot2::autoplot(bench1e6)
```

```{r}
#| label: tab-bench1e6
reorderMicrobenchmarkResults(bench1e6)
```


```{r}
#| label: bench-1e8
#| cache: true
D <- data.table::fread("measurements1e8.csv", stringsAsFactors = TRUE)

bench1e8 <- microbenchmark::microbenchmark(
    aggregate = aggregate(measurement ~ state, data = D, FUN = sum_stats_vec),
    split_lapply = split_lapply(D),
    array2DF_by = array2DF(by(D$measurement, D$state, sum_stats_vec)),
    raw_by = by(D$measurement, D$state, sum_stats_vec),
    docall_tapply = do.call("rbind", tapply(D$measurement, D$state, sum_stats_vec)),
    sapply_tapply = sapply(tapply(D$measurement, D$state, sum_stats_vec), rbind),
    reduce = reduce(D),
    Rfast = Rfast(D),
    dplyr = D |> dplyr::group_by(state) |> dplyr::summarise(sum_stats_tibble(measurement)) |> dplyr::ungroup(),
    datatable = D[, .(sum_stats_list(measurement)), by = state],
    times = 10
)
```

```{r}
#| label: fig-bench1e8
ggplot2::autoplot(bench1e8)
```

```{r}
#| label: tab-bench1e8
reorderMicrobenchmarkResults(bench1e8)
```