---
title: "vellumverse #7: A grammar of graphics that compiles"
author:
- name: David Schoch
  orcid: 0000-0003-2952-4812
date: '2026-07-28'
categories:
- R
- data visualization
- vellumverse
---

Five posts of backend, and we finally climb a layer. `vellum` gave us
primitives, viewports, a retained tree, and a read-back table, and it stopped
short of scales, geoms, and facets on purpose. Those live in `vellumplot`, the
grammar of graphics that compiles down into a `vellum` scene, and that is where
the rest of this series lives.

If you know `ggplot2`, the first plot looks familiar.

```{r}
library(vellumplot)

vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp) |>
  scale_color_continuous()
```

`vplot()` starts from a data frame, `mark_point()` adds a layer, and encodings
like `x = wt` and `color = hp` are tidy-eval expressions evaluated against the
data. The pipe replaces `+`, which fits R's native `|>` and reads left to right.

The difference is underneath, and it is the one claim of this post: building that
plot drew nothing. `vplot(...) |> mark_point(...)` returns a *spec*, a plain S7
object recording the data, the layers, the scales, the facet, the coord, and the
theme. It is a value, not a side effect, and you can look at it before you ever
render it.

```{r}
p <- vplot(mtcars) |> mark_point(x = wt, y = mpg, color = hp)
summary(p)
```

`ggplot2` does something similar, since a `ggplot` object is also built before it
prints, so this is a matter of degree rather than a brand-new idea. But
`vellumplot` leans on it much harder. The spec is designed to be inspected and
serialized: it is the answer to "what will this draw?" available without drawing.
A few consequences follow that I use constantly.

A plot is an ordinary value, so you can treat it like one: build it inside a
function and return it, keep a list of specs and render them in a loop, or pipe
another mark onto a plot someone handed you. You can even save the spec to disk
and rebuild the exact figure next month. None of these need a device open,
because none of them draw.

It also makes plots testable. You can assert on the spec, on how many layers it
carries or which scale trained where, without rendering a pixel and diffing
images. For a package with this many marks, that is the difference between a test
suite that runs in milliseconds and one that shells out to a rasterizer.

Nothing draws until you print the spec or call `render_plot()`. That is when a
real compiler runs: it takes the spec and turns it into the `vellum` scene from
the last few posts, then hands that scene to the backend. `vellumplot` never
touches a graphics device itself. It produces a scene and steps back.

```{r}
#| eval: false
render_plot(p, "cars.png")   # compile the spec, then render the scene
```

So the mental model is two clean halves. You *describe* a plot as data, and later
a compiler *turns that description into a scene*. Everything good about the
grammar layer, scales that span layers, faceting that does not redraw, and one
plot rendering to three formats, comes from keeping those two halves apart.

Next post I open the compiler and walk the stages it runs, from tidy-eval
expressions to positioned `vellum` grobs.
