---
title: "vellumverse #13: Coming from ggplot2, and plugging in your own"
author:
- name: David Schoch
  orcid: 0000-0003-2952-4812
date: '2026-08-07'
categories:
- R
- data visualization
- vellumverse
draft: true
---

Thirteen posts in, here is the practical send-off. If you think in `ggplot2`, you
already think in this grammar, and the gap between them is smaller than the length
of this series makes it look. Four things differ, and after those it is mostly
renaming.

The pipe replaces the plus, so a plot is an ordinary pipeline whose first argument
is the plot. There is no `aes()`: aesthetics are named arguments on the mark,
where a bare column name maps through a scale and a literal is a constant, so
`color = hp` draws a legend and `color = "red"` does not. Every `geom_*` becomes a
`mark_*` with the same job. And, as the whole middle of this series laboured to
show, a plot is a spec you can inspect, not a picture that has already drawn.

Most of what you know transfers by find-and-replace:

| ggplot2 | vellumplot |
|---|---|
| `ggplot(df, aes(x, y))` | `vplot(df)` then `mark_*(x = x, y = y)` |
| `aes(color = z)` | `color = z` on the mark |
| `+` | `\|>` |
| `geom_point()`, `geom_smooth()` | `mark_point()`, `mark_smooth()` |
| `geom_col()`, `geom_histogram()` | `mark_bar()`, `mark_histogram()` |
| `facet_wrap(~z)`, `facet_grid(a ~ b)` | same names, same formulas |
| `scale_*_manual(values = )` | `scale_*_manual(values = )` |
| `theme_minimal()`, `labs(title = )` | same |
| `ggsave("p.png")` | `render_plot(p, "p.png")` |

So a scatter with a fitted line is the plot you would expect, with `+` swapped for
`|>` and each mark carrying its own encodings rather than inheriting a shared
`aes()`:

```{r}
#| eval: false
library(vellumverse)
vplot(mtcars) |>
  mark_point(x = wt, y = mpg) |>
  mark_smooth(x = wt, y = mpg, method = "lm")
```

A couple of habits do not carry over, on purpose. There is no global `aes()`
shared across layers, so each mark states its own encodings: more typing on a
multi-layer plot, and one fewer set of inheritance rules to hold in your head. In
return you get two things `ggplot2` cannot easily give you. `summary(p)` prints a
plot's structure without drawing it, and `as_widget(p)` makes the same plot
interactive with no rewrite.

That covers using the ecosystem. The last thing worth showing is how to extend it,
because the seams the three layers agree on are open to you too. The deepest one is
a single S7 generic, `as_vellum_scene()`, that every consumer calls to turn
something into a scene. Register a method for your own object and it renders and
goes interactive through the exact same machinery as a `vplot`.

```{r}
#| eval: false
library(S7)

my_plot <- function(data) structure(list(data = data), class = "my_plot")

method(as_vellum_scene, new_S3_class("my_plot")) <- function(x, ...) {
  vl_scene(4, 3, bg = "white") |>
    push(vl_viewport(xscale = range(x$data$x), yscale = range(x$data$y))) |>
    draw(points_grob(vl_unit(x$data$x, "native"), vl_unit(x$data$y, "native"),
                     gp = vl_gpar(fill = "#6b4f2c")))
}

p <- my_plot(data.frame(x = 1:10, y = (1:10)^2))
render(p, "custom.png")   # works
as_widget(p)              # also works
```

That one method is the whole price of admission. `render()`, `render_plot()`, and
`as_widget()` never learn what `my_plot` is; they see only the scene it compiles
to, and everything downstream, the three output formats, the tooltips, the linked
selection, comes along for free. It is the clearest case for why the ecosystem is
split into layers at all: a new plot type built at the top inherits every
capability built underneath.

That is the series. It began as an idle question, whether R could lean on a Rust
engine for graphics at all. Poking at `grid` turned up the one constraint behind
most of its friction, and from there we built up a Rust scene graph, a retained
editable tree, a compiler, and a grammar that renders one specification to a file
or a widget without rewriting a line. If you want to try it, install `vellumverse`
and start with a `vplot()`. Everything else in these thirteen posts is what
happens after you print it.
