---
title: "vellumverse #9: Marks and scales: the parts you actually touch"
author:
- name: David Schoch
  orcid: 0000-0003-2952-4812
date: '2026-07-31'
categories:
- R
- data visualization
- vellumverse
---

The last two posts were about machinery. This one is the opposite: the small
surface you actually touch to make real plots, and how little of it there is.

Start with the grammar rule that does the most work. Inside a `mark_*()`, a bare
column name is an *encoding*, mapped through a scale, and a scalar is a
*constant*. So `color = hp` colours points by the `hp` column and draws a legend,
while `color = "red"` paints every point red and draws none. That one
distinction, expression versus value, is most of what you need to read any
`vellumplot` code.

```{r}
library(vellumplot)
vplot(mtcars) |>
  mark_point(x = wt, y = mpg, color = hp, size = 3)
```

`color = hp` maps; `size = 3` is a constant marker. Same function call, two kinds
of argument.

The marks come in a few families, and the names are boring on purpose. Points and
lines are `mark_point()`, `mark_line()`, `mark_step()`, `mark_rule()`. Bars and
areas are `mark_bar()` (heights, or a row count when you omit `y`), `mark_area()`,
and `mark_ribbon()`. Intervals are `mark_errorbar()`, `mark_linerange()`, and
`mark_segment()`. Tiles and bins are `mark_tile()`, `mark_hex()`, and
`mark_bin2d()`. Then there is text, images, and the statistical marks
(`mark_smooth()`, `mark_boxplot()`, `mark_contour()`) that compute something
before they draw. On a normal day you reach for maybe six of these.

Layering is where scales earn their keep. Pipe two marks and they share the
trained axes automatically, because (as the last post showed) scales train across
every layer before anything draws. A scatter with a fitted line is just two marks
on one panel.

```{r}
peng <- na.omit(datasets::penguins)
vplot(peng) |>
  mark_point(x = bill_len, y = body_mass, color = species, alpha = 0.5) |>
  mark_smooth(x = bill_len, y = body_mass, method = "lm", se = TRUE) |>
  labs(title = "Linear smooth with confidence band",
       x = "bill length (mm)", y = "body mass (g)")
```

The `color = species` encoding trains one discrete colour scale that both the
points and the legend read from. `mark_smooth()` fits its linear model in the
compiler's stat stage and draws the band on top. `labs()` sets the titles.
Nothing here aligned an axis or placed a legend by hand.

Scales are how you override the defaults when the automatic choice is wrong. Each
aesthetic has a family of `scale_*()` functions (`scale_color_continuous()`,
`scale_fill_gradient(low =, high =)`, `scale_fill_continuous(palette =)`, and so
on). Add one to the pipe and it replaces the trained default for that aesthetic.
A correlation matrix wants a diverging fill, so you say so.

```{r}
cm <- cor(mtcars[, c("mpg", "disp", "hp", "drat", "wt", "qsec")])
hm <- expand.grid(v1 = rownames(cm), v2 = colnames(cm))
hm$cor <- as.vector(cm)

vplot(hm) |>
  mark_tile(x = v1, y = v2, fill = cor) |>
  scale_fill_gradient(low = "#b2182b", high = "#2166ac") |>
  labs(title = "Correlation matrix", x = NULL, y = NULL)
```

`mark_tile()` draws a rectangle per `(x, y)` cell, `fill = cor` maps the
correlation, and the explicit `scale_fill_gradient()` swaps the default
continuous fill for a red-to-blue diverging ramp. A complete figure in five
lines.

That is genuinely most of the day-to-day API: `vplot()`, a mark or two, an
encoding for each thing you want mapped, a `scale_*()` when the default is wrong,
and `labs()`. Facets, coords, and themes are waiting when you need them, and the
reference has the full catalogue, but the handful above gets real work done.

We have built plots now. The next post cashes in the promise the series has been
saving toward: this spec compiles to one `vellum` scene, and that single scene
renders to a raster, a vector, or an interactive widget without you rebuilding a
thing.
