---
title: "vellumverse #5: The one superpower of keeping the scene around"
author:
- name: David Schoch
  orcid: 0000-0003-2952-4812
date: '2026-07-24'
categories:
- R
- data visualization
- vellumverse
draft: true
---

If you take one idea from this series, make it this one. Everything interactive
that comes later is a consequence of it, and it is the cleanest thing that falls
out of putting the model in Rust.

Most drawing APIs, `grid` included, are immediate mode: a call paints and
forgets. The pixels exist afterwards; the thing that drew them does not. `vellum`
is retained mode. The scene you build stays as a tree of nodes until you render
it, and that tree is still there to be questioned and rewritten afterwards.

Addressability starts with a name. Every grob takes an optional `name`, and a
named node is one you can find again. Here is a bar chart of seven named
rectangles.

```{r}
library(vellum)

vals <- c(3, 5, 2, 8, 6, 4, 7)
n <- length(vals); y0 <- 0.18
base <- "#c9a874"; ink <- "#3a2f1e"

scene <- vl_scene(7, 3.2, bg = "#faf5ea")
for (i in seq_len(n)) {
  h <- 0.62 * vals[i] / max(vals)
  scene <- scene |>
    draw(rect_grob(x = (i - 0.5) / n, y = y0 + h / 2,
                   width = 0.7 / n, height = h, name = paste0("bar", i),
                   gp = vl_gpar(fill = base, col = NA)))
}
scene <- scene |>
  draw(segments_grob(x0 = 0.02, y0 = y0, x1 = 0.98, y1 = y0,
                     gp = vl_gpar(col = ink, lwd = 1.5)))

node_names(scene)
```

`node_names()` lists what is addressable, in paint order. `get_node()` hands one
back so you can read its current state, which is ordinary R data rather than an
internal handle.

Editing works the way you would hope. To highlight the tallest bar, find its name
and rewrite that one node's fill. No loop over the others, no rebuild of the
scene.

```{r}
top <- paste0("bar", which.max(vals))
highlighted <- edit_node(scene, top, gp = vl_gpar(fill = "#c1121f", col = NA))
highlighted
```

`edit_node()` returns a *new* scene with the change applied; the original `scene`
still shows every bar in the base colour. That copy-on-modify guarantee is what
makes a scene safe to pass around: you can derive as many variants as you like
from one base graph and none of them disturbs the others. The new tree shares
structure with the old one underneath, so deriving a variant stays cheap even for
a large scene.

This is exactly how a host implements hover and selection later. It keeps one
scene, and on each interaction re-derives it with the touched node restyled, then
re-renders. Flag that node's viewport as a repaint boundary and only the changed
subtree re-rasterises, so the update is cheap.

Retention answers the inverse question too. `hit_test()` takes a point and tells
you which node is drawn on top there.

```{r}
#| eval: false
hit_test(scene, x = 0.5, y = 0.4)   # the bar under that point, by name
```

`grid` offered only `grid.locator()`, a single interactive click. Because a
retained scene can be compiled into a colour pick-buffer, `hit_test()` is exact
with respect to geometry, clipping, and paint order, and you can call it
programmatically as often as you want. That is enough to route a click on the
canvas back to the datum that drew the mark.

Naming, editing, and hit-testing are three small functions, and together they are
the difference between a picture and a thing you can interrogate. Later in the
series, when a widget adds tooltips and brushing with no server involved, it will
need no new drawing machinery. It will just be standing on this.

Next: if a scene is a tree you can read back, what exactly do you get when you
read it? The answer is a plain table, and it turns out to be the contract the
entire ecosystem is built on.
