---
title: "vellumverse #3: vellum, where a scene is a value you can hold"
author:
- name: David Schoch
  orcid: 0000-0003-2952-4812
date: '2026-07-21'
categories:
- R
- data visualization
- vellumverse
---

[Two posts back](https://blog.schochastics.net/posts/2026-07-17_two-graphics-systems/index.html) I described the experiment this all started as. [In the last post](https://blog.schochastics.net/posts/2026-07-20_inside-a-graphics-engine/index.html), I
named the seven parts a graphics engine is built from. Now I get to show you `vellum`,
which builds those seven parts in Rust and hands R a small declarative API on top.
This post makes one claim: in `vellum` a scene is an ordinary value you construct,
inspect, and render, not a side effect fired at a device.

Start with the smallest thing that draws.

```{r}
library(vellum)

vl_scene(width = 6, height = 3, bg = "white") |>
  draw(circle_grob(x = 0.2, y = 0.5, r = 0.28,
                   gp = vl_gpar(fill = "#f7c948", col = NA))) |>
  draw(text_grob("vellum", x = 0.6, y = 0.5,
                 gp = vl_gpar(fontsize = 44, col = "#1b2a4a", fontface = "bold")))
```

`vl_scene()` fixes the page: six inches by three, white background. Each `draw()`
appends a grob, one of a couple dozen primitives such as `circle_grob()` and
`text_grob()`. Coordinates default to `"npc"`, normalised parent coordinates,
where `(0, 0)` is the bottom-left of the region and `(1, 1)` the top-right. A
grob on its own draws nothing; it is a value. The scene is a value too, and
printing it is what renders it.

## Viewports, and data coordinates

A flat page is not much of an engine. The tree comes from viewports: `push()`
descends into a `vl_viewport()`, `draw()` adds a grob at the current level, and
`pop()` climbs back up. A viewport carries its own coordinate system, so once you
give it an `xscale` and a `yscale`, `"native"` units place raw data values
straight onto the region.

```{r}
set.seed(1)
x <- runif(60, 0, 10)
y <- 1.8 * x + rnorm(60, 0, 4)

vl_scene(5, 3.2, bg = "white") |>
  push(vl_viewport(width = 0.82, height = 0.82,
                   xscale = c(0, 10), yscale = range(pretty(y)))) |>
  draw(rect_grob(gp = vl_gpar(fill = "#f4f6f8", col = "#cfd8dc"))) |>
  draw(points_grob(vl_unit(x, "native"), vl_unit(y, "native"),
                   size = vl_unit(3.2, "mm"),
                   gp = vl_gpar(fill = "#3a7bd5", col = "#1b2a4a"))) |>
  pop()
```

The panel rectangle is drawn in `"npc"`, so it fills its viewport at any size.
The points sit in `"native"`, so they follow the data scales. The marker size in
millimetres stays millimetres no matter how the panel is resized. Mixing those
coordinate systems in one scene is the whole trick behind panels, insets, and
eventually facets: push a region, draw in its local coordinates, pop back out.

If you know `grid`, none of this is new. `vl_scene()` is `grid.newpage()`,
`push()`/`pop()` are `pushViewport()`/`popViewport()`, and `vl_unit(1, "native")`
is `unit(1, "native")`. What you feel immediately is that no global device or
viewport stack is being mutated behind your back. The pipe is the tree, and the
scene at the end is a plain R value you can assign to a variable and branch from.

## One value, three formats

Because the scene is a value, rendering is a separate step you call when you want
a file. The extension picks the backend.

```{r}
#| eval: false
s <- vl_scene(4, 3) |>
  draw(circle_grob(r = 0.3, gp = vl_gpar(fill = "tomato", col = NA)))

render(s, "out.png")   # raster
render(s, "out.svg")   # vector
render(s, "out.pdf")   # vector
```

Same geometry across all three, and the raster and PDF output are byte-stable, so
a figure is reproducible and snapshot-testable. Hold onto that split between the
scene and its rendering, because it is the hinge the whole ecosystem turns on
later in the series.

I have been saying "in Rust" and walking straight past it. The next post stops
and asks why: what actually runs in Rust, what it buys, and what it costs.
