---
title: "vellumverse #10: One scene, three outputs"
author:
- name: David Schoch
  orcid: 0000-0003-2952-4812
date: '2026-08-03'
categories:
- R
- data visualization
- vellumverse
draft: true
---

This is the post the whole series has been walking toward. Everything so far, the
retained scene, the contract, the compiler, exists to make one move possible: you
describe a plot once, and where it ends up is a decision you make at the end, not
a fork you commit to at the start.

Most plotting workflows re-specify the plot per destination. One code path for the
PNG in a report, a different one for the vector figure a journal wants, and a
third library entirely when the thing has to be interactive on the web. Those
three drift over time, because they are three plots. Here the specification and
the destination are separate concerns, joined by a single seam.

Build the spec once. Notice that `tooltip` and `data_id` are declared right here,
on the mark, even though a static render ignores them.

```{r}
library(vellumverse)

df <- data.frame(wt = mtcars$wt, mpg = mtcars$mpg, model = rownames(mtcars))

p <- vplot(df) |>
  mark_point(x = wt, y = mpg, color = mpg, tooltip = model, data_id = model) |>
  scale_color_continuous() |>
  labs(title = "Fuel economy")
```

Nothing has drawn. `p` is a description. Now send it three places.

Raster, for the report:

```{r}
#| eval: false
render_plot(p, "fuel.png")
render_plot(p, "fuel.png", dpi = 300)   # override the authored resolution
```

Vector, for the journal:

```{r}
#| eval: false
render_plot(p, "fuel.svg")
render_plot(p, "fuel.pdf")
```

Interactive, for the web:

```{r}
#| eval: false
as_widget(p)
```

Same `p` every time. Under the hood, `render_plot()` and `as_widget()` both call
one generic, `as_vellum_scene()`, which compiles the spec into the `vellum` scene
from the middle of this series. Render walks that scene to a raster or vector
file; the widget emits the scene's SVG plus the per-element table from the
contract post and wires up behaviour. One spec, one compiled scene, three roads
out.

The payoff is more than fewer keystrokes. Because there is exactly one scene
behind all three outputs, they cannot drift: the PNG in your paper and the widget
on your blog are guaranteed to show the same plot, because they *are* the same
scene, laid out once. Point positions, text metrics, and legend placement are
computed a single time and then walked for each backend. The raster and the
vector are one plot, not two approximations of it.

The decision genuinely defers, too. You can write analysis code that returns a
spec and hand it off, letting the calling context decide how to realise it: a
Quarto chunk prints it, a Shiny app makes it a widget, a batch job writes a
hundred PDFs. The function that built the plot never knew which.

This works one layer down as well, without the grammar. A bare `vellum` scene
built from grobs travels the same three roads, because `render()` and
`as_widget()` accept a scene directly. The gallery's "one scene, three formats"
entry does exactly that with a hand-built logo: one scene value, `render()` to
PNG, SVG, and PDF, pixel-for-pixel agreement. The grammar in `vellumplot` is, in
the end, a convenient way to produce one of these scenes.

That is the thing the experiment was really testing, and it held up. The next
post takes the third road seriously: what the interactive widget actually does,
and why it needs no server and no new drawing code at all.
