---
title: "vellumverse #11: Interactivity with no server"
author:
- name: David Schoch
  orcid: 0000-0003-2952-4812
date: '2026-08-04'
categories:
- R
- data visualization
- vellumverse
---

Several posts back, in the retained-mode post, I asked you to hold onto one idea:
because a `vellum` scene is retained, you can reach back into it and edit any node
by name. This is the post that spends it. Everything the interactive widget does is built on that one
property, and it needs no server and no new drawing code.

There is one verb. `as_widget()` turns a `vellumplot` plot (or a bare `vellum`
scene) into a self-contained HTML widget that behaves the same in the RStudio
viewer, a Quarto document, and a Shiny app. Every interaction runs in the
browser, with no server and no round-trip.

The interactivity is declared in the grammar rather than bolted on afterwards. A
mark's `tooltip`, `data_id`, and `hover_group` are just more arguments, and they
flow through the scene as the per-element metadata from the contract post. Hover a
point and it shows its tooltip and dims the rest; `data_id` is the join key and
`tooltip` is what the box says.

```{r}
#| eval: false
library(vellumverse)
df <- data.frame(wt = mtcars$wt, mpg = mtcars$mpg,
                 model = rownames(mtcars), cyl = factor(mtcars$cyl))

vplot(df) |>
  mark_point(x = wt, y = mpg, color = cyl, tooltip = model, data_id = model) |>
  as_widget()
```

That single call gets you hover tooltips, click-to-select (every mark sharing a
`data_id` toggles together), rectangular brush-select, and pan/zoom by wheel,
drag, touch, or arrow key, plus a toolbar with SVG and PNG export. Accessibility is
on by default: you can Tab into the chart, arrow between marks with each announced
through a live region, and select with Enter, and a visually-hidden data table
backs it all for screen readers.

Here is the connection I want to make explicit. None of that required new drawing
machinery. Recall how editing worked back in the retained-mode post: keep one
scene, and on an interaction re-derive it with the touched node restyled, then
re-render only the changed subtree. Hover and selection are exactly that, run in
the browser. The widget knows which element you touched because every keyable mark
carries a `key`, and it knows where each element sits because `scene_model()`
handed it the device-pixel bounding box. Retained mode gave it addressability; the
contract gave it the table. The widget is a thin client over the two.

Because identity lives in the data keys rather than in pixels, linking views is
nearly free. Two widgets that share a `group` highlight the same keys in each
other, client-side, with no Shiny and no crosstalk:

```{r}
#| eval: false
p <- vplot(df) |> mark_point(x = wt, y = mpg, data_id = model, hover_group = cyl)
htmltools::tagList(
  as_widget(p, group = "cars", width = 320, height = 240),
  as_widget(p, group = "cars", width = 320, height = 240)
)
```

Select in one, and the same cars light up in the other. To link against plotly,
leaflet, or a DT table instead, pass a `crosstalk::SharedData` to `crosstalk =`
and the same keys drive those too. And when you genuinely do need the server,
`vellumwidgetOutput()` and `renderVellumwidget()` report the selection, clicks,
hovers, and brush back to Shiny as reactive inputs, keyed the same way.

The thing to take from this is how little had to be added. The backend was already
retained and addressable, and the scene already carried keys and geometry, so
"make it interactive" turned out to be a browser reading a table and restyling
nodes. No server was ever in the picture.

The next post pushes the opposite direction, toward scale: what to do when the
data has more points than the screen has pixels, and drawing every one is the
wrong move.
